---
title: "DDIWAS R package tutorial"
author: "Patrick Wu"
date: "10/17/2020"
output: 
  html_document:
    theme: paper
vignette: >
  %\VignetteIndexEntry{DDIWAS R package tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Description
CSV files in this tutorial come from the [vignette](https://pwatrick.github.io/ddiwas/articles/extract_ehr_data.html) on how to extract EHR data for DDIWAS.  

* This tutorial shows how to process and analyze electronic health record (EHR) data extracted from a relational database organized using the [OHDSI/OMOP Common Data Model](http://archive.today/2020.10.14-102308/https://ohdsi.github.io/TheBookOfOhdsi/CommonDataModel.html).  
* Please submit questions/comments/feedback to wei-qi.wei(AT)vumc.org.  
  
## Import packages
Load required packages
```{r}
suppressPackageStartupMessages({
  library(tidyverse);
  library(broom); 
  library(ddiwas); 
  library(glue); 
  library(lubridate)})
set.seed(1)
```
  
## Map drug allergies
```{r, eval = FALSE}
ehr_drug_allergies <- read_csv("ddi_adr_data.csv", col_types = cols())
names(ehr_drug_allergies) <- tolower(names(ehr_drug_allergies))
drug_allergies <- ddiwas::map_drug_allergies(ehr_drug_allergies = ehr_drug_allergies)
```
  
Example **input** `ddi_adr_data.csv` dataframe  
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~NOTE_TEXT, ~FIRST_ADR_DATE, ~START_DATE, ~END_DATE,
  1, 'gemfibrozil', '2011-09-01', '2010-10-28', '2011-10-28',
  2, 'No known allergies', '2012-09-15', '2012-06-24', '2013-04-13'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Example **output** `drug_allergies` dataframe  
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~drug_allergy,
  1, "gemfibrozil"
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
## Process covariates
```{r, eval = FALSE}
df_drug_exposures <- read_csv("ddi_drug_exposures.csv", col_types = cols())
names(df_drug_exposures) <- c("person_id", "concept_id")

covariates <- read_csv("ddi_covariates.csv", col_types = cols())
names(covariates) <- tolower(names(covariates))
covariates <- ddiwas::process_covariates(covariates, df_drug_exposures)
```
  
Example **input** dataframes  

* `ddi_drug_exposures.csv`  
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~DRUG_CONCEPT_ID,
  1, 1308877, 
  2, 43014173
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
* `ddi_covariates.csv`  
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~START_DATE, ~END_DATE, ~GENDER, ~DOB, ~RACE,
  1, 1, as.Date('2010-10-28'), as.Date('2011-10-28'), 'F', as.Date('1950-05-06'), 'W', 
  2, 0, as.Date('2012-06-24'), as.Date('2013-04-13'), 'M', as.Date('1965-12-12'), 'W'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Example **output** `covariates` dataframe
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~group_label, ~start_date, ~end_date, ~gender, ~dob, ~race, ~age, ~obs_length, ~num_drug_exposures, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~is_m, ~is_w,
  1, 1, as.Date('2010-10-28'), as.Date('2011-10-28'), 'F', as.Date('1950-05-06'), 'W', 60.5, 365, 9, 0.590, 0.997, 0.0552, 0, 1,
  2, 0, as.Date('2012-06-24'), as.Date('2013-04-13'), 'M', as.Date('1965-12-12'), 'W', 46.5, 293, 12, 0.396, 0.801, 0.0759, 1, 1
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
## Require allergy list entry during observation period
```{r, eval = FALSE}
covariates <- ddiwas::require_drug_allergies(ehr_drug_allergies, covariates)
```
  
Example **input** dataframes  

* `ehr_drug_allergies`  
```{r echo=FALSE}
tribble(
  ~person_id, ~note_text, ~first_adr_date, ~start_date, ~end_date,
  1, 'gemfibrozil', '2011-09-01', '2010-10-28', '2011-10-28',
  2, 'No known allergies', '2012-09-15', '2012-06-24', '2013-04-13'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```

* `covariates`  
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~group_label, ~start_date, ~end_date, ~gender, ~dob, ~race, ~age, ~obs_length, ~num_drug_exposures, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~is_m, ~is_w,
  1, 1, as.Date('2010-10-28'), as.Date('2011-10-28'), 'F', as.Date('1950-05-06'), 'W', 60.5, 365, 9, 0.590, 0.997, 0.0552, 0, 1,
  2, 0, as.Date('2012-06-24'), as.Date('2013-04-13'), 'M', as.Date('1965-12-12'), 'W', 46.5, 293, 12, 0.396, 0.801, 0.0759, 1, 1
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
Example **output** `covariates` dataframe after requiring allergy list entry
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~is_m, ~is_w, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~group_label,
  1, 0, 1, 0.590, 0.997, 0.0552, 1, 
  2, 1, 1, 0.396, 0.801, 0.0759, 0
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
## Create regression table
```{r, eval = FALSE}
regression_table <- ddiwas::create_regression_table(drug_allergies, covariates)
```
  
Example **input** dataframes.  

* `drug_allergies`  
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~drug_allergy,
  1, "gemfibrozil"
) %>% DT::datatable(options = list(scrollX = TRUE))
```

* `covariates`  
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~is_m, ~is_w, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~group_label,
  1, 0, 1, 0.590, 0.997, 0.0552, 1, 
  2, 1, 1, 0.396, 0.801, 0.0759, 0
) %>% DT::datatable(options = list(scrollX = TRUE))
```

Example **output** `regression_table` dataframe
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~is_m, ~is_w, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~group_label, ~gemfibrozil, ~lisinopril, 
  1, 0, 1, 0.590, 0.997, 0.0552, 1, 1, 0, 
  2, 1, 1, 0.396, 0.801, 0.0759, 0, 0, 0
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
## Run regression
```{r, eval = FALSE}
results <- ddiwas::ddiwas_regression(regression_table, rare_events = TRUE)
```
  
### Add `object_drug` column. Rearrange and rename columns.  
```{r, eval = FALSE}
results$object_drug <- "simvastatin"
results <- results %>% 
  select(object_drug, drug, coef, se, pval, or, nA, nB, nC, nD)
names(results) <- c("object_drug", "potential_precipitant_drug", "coef", "se", "pval", "or", "nA", "nB", "nC", "nD") 
```

Example **input** `regression_table` dataframe.  
```{r echo=FALSE}
dplyr::tribble(
  ~person_id, ~is_m, ~is_w, ~age_n, ~obs_length_n, ~num_drug_exposures_n, ~group_label, ~gemfibrozil, ~lisinopril, 
  1, 0, 1, 0.590, 0.997, 0.0552, 1, 1, 0, 
  2, 1, 1, 0.396, 0.801, 0.0759, 0, 0, 0
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
Example **output** `results` dataframe.  
```{r echo=FALSE}
dplyr::tribble(
  ~object_drug, ~potential_precipitant_drug, ~coef, ~se, ~pval, ~or, ~nA, ~nB, ~nC, ~nD,
  "simvastatin", "gemfibrozil", 1.98, 0.32, 1.14E-09, 7.21, 12, 49, 2802, 85824,
  "simvastatin", "lisinopril", 0.66, 0.14, 2.21E-06, 1.93, 57, 853, 2757, 85020
) %>% DT::datatable(options = list(scrollX = TRUE))
```
  
## Packages used to create document
```{r}
sessionInfo()
```
