---
title: "Code to extract electronic health record (EHR) data from relational databases for Drug-Drug Interaction Wide Association Studies (DDIWAS)"
author: "Patrick Wu"
date: "10/17/2020"
output: 
  html_document:
    theme: paper
vignette: >
  %\VignetteIndexEntry{Code to extract electronic health record (EHR) data from relational databases for Drug-Drug Interaction Wide Association Studies (DDIWAS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(tidyverse)
```
## Description
To use the output CSV files from this example document, see [vignette](https://pwatrick.github.io/ddiwas/articles/ddiwas_r_package_tutorial.html) on using the DDIWAS R package.  

* This document provides example code to extract electronic health record (EHR) data organized using the [OHDSI/OMOP Common Data Model](http://archive.today/2020.10.14-102308/https://ohdsi.github.io/TheBookOfOhdsi/CommonDataModel.html).  
* There may be some institution-specific data dependencies that require minor modifications if you use the code outside of the Synthetic Derivative at Vanderbilt University Medical Center. For example, allergy list data (see `ddi2` table) may be located not in the `NOTE` table, but rather the `NOTE_NLP` table. See [discussion](http://archive.today/2020.10.03-114046/https://github.com/OHDSI/CommonDataModel/issues/85) on `NOTE_NLP` tables.  
* The code is written to extract data from a IBM Netezza relational database.  
* Please submit questions/comments/feedback to wei-qi.wei(AT)vumc.org.  

## Extract EHR data to identify `simvastatin` drug interactions using DDIWAS.  
### Identify patients exposed to `simvastatin` using generic and brand names. 
```{sql,eval = FALSE}
create table ddi1 as 
(select t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id
from (select * from drug_exposure) t1
where lower(t1.drug_source_value) like '%simvastatin%'
group by t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id)
union
(select t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id
from (select * from drug_exposure) t1
where lower(t1.drug_source_value) like '%zocor%'
group by t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id)
union
(select t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id
from (select * from drug_exposure) t1
where lower(t1.drug_source_value) like '%flolipid%'
group by t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id)
union
(select t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id
from (select * from drug_exposure) t1
where lower(t1.drug_source_value) like '%vytorin%'
group by t1.drug_exposure_id, t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_source_concept_id);
```
  
Example `ddi1` table
```{r echo=FALSE}
ddi1_df <- dplyr::tribble(
  ~DRUG_EXPOSURE_ID, ~PERSON_ID, ~DRUG_EXPOSURE_START_DATE, ~DRUG_SOURCE_VALUE, ~DRUG_SOURCE_CONCEPT_ID,
  1, 1, '2010-03-21', 'Zocor', 1539406,
  2, 1, '2011-03-17', 'Zocor', 1539406, 
  3, 1, '2011-10-28', 'Zocor', 1539406,
  4, 2, '2012-06-24', 'Simvastatin', 1539411, 
  5, 2, '2013-04-13', 'Simvastatin', 1539411, 
)

DT::datatable(ddi1_df, options = list(scrollX = TRUE))
```

### Define observation period
Identify patients with `simvastatin` listed in their allergy list  
```{sql,eval = FALSE}
create table ddi2 as
(select t1.person_id, t1.note_date, t1.note_text, t1.note_source_value
from (select * from note) t1
where lower(note_source_value) like '%allerg%'
  and lower(note_text) like '%simvastatin%')
union
(select t1.person_id, t1.note_date, t1.note_text, t1.note_source_value
from (select * from note) t1
where lower(note_source_value) like '%allerg%'
  and lower(note_text) like '%zocor%')
union
(select t1.person_id, t1.note_date, t1.note_text, t1.note_source_value
from (select * from note) t1
where lower(note_source_value) like '%allerg%'
  and lower(note_text) like '%flolipid%')
union
(select t1.person_id, t1.note_date, t1.note_text, t1.note_source_value
from (select * from note) t1
where lower(note_source_value) like '%allerg%'
  and lower(note_text) like '%vytorin%');
```
  
Example `ddi2` table
```{r echo=FALSE}
ddi2_df <- dplyr::tribble(
  ~PERSON_ID, ~NOTE_DATE, ~NOTE_TEXT, ~NOTE_SOURCE_VALUE,
  1, '2011-10-28', 'Zocor (muscle cramps)', 'ALLERGIES',
  1, '2012-03-07', 'Zocor (muscle cramps)', 'ALLERGIES'
)

DT::datatable(ddi2_df, options = list(scrollX = TRUE))
```

For each patient in `ddi2`, get first date that simvastatin was listed in their allergy list
```{sql, eval = FALSE}
create table ddi3 as
select t1.person_id, min(t1.note_date) as case_end_date
from (select * from ddi2) t1
group by t1.person_id;
```
  
Example `ddi3` table
```{r echo=FALSE}
ddi3_df <- dplyr::tribble(
  ~PERSON_ID, ~CASE_END_DATE,
  1, '2011-10-28'
)

DT::datatable(ddi3_df, options = list(scrollX = TRUE))
```

For each patient in `ddi1`, get first date of simvastatin exposure
```{sql, eval = FALSE}
create table ddi4 as 
select t1.person_id, min(t1.drug_exposure_start_date) as first_date_simvastatin_exposure
from (select * from ddi1) t1
group by t1.person_id;
```
  
Example `ddi4` table
```{r echo=FALSE}
ddi4_df <- dplyr::tribble(
  ~PERSON_ID, ~FIRST_DATE_SIMVASTATIN_EXPOSURE,
  1, '2010-03-21', 
  2, '2012-06-24'
)

DT::datatable(ddi4_df, options = list(scrollX = TRUE))
```

For each patient in `ddi1`, get last date of simvastatin exposure
```{sql, eval = FALSE}
create table ddi5 as 
select t1.person_id, max(t1.drug_exposure_start_date) as last_date_simvastatin_exposure
from (select * from ddi1) t1
group by t1.person_id;
```
  
Example `ddi5` table
```{r echo=FALSE}
ddi5_df <- dplyr::tribble(
  ~PERSON_ID, ~LAST_DATE_SIMVASTATIN_EXPOSURE,
  1, '2011-10-28',
  2, '2013-04-13'
)

DT::datatable(ddi5_df, options = list(scrollX = TRUE))
```

Combine tables `ddi4` and `ddi5` to get first and last dates of simvastatin exposure in one table
```{sql, eval = FALSE}
create table ddi6 as 
select t1.person_id, t1.first_date_simvastatin_exposure, t2.last_date_simvastatin_exposure
from (select * from ddi4) t1, (select * from ddi5) t2
where t1.person_id = t2.person_id
group by t1.person_id, t1.first_date_simvastatin_exposure, t2.last_date_simvastatin_exposure;
```
  
Example `ddi6` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~FIRST_DATE_SIMVASTATIN_EXPOSURE, ~LAST_DATE_SIMVASTATIN_EXPOSURE,
  1, '2010-03-21', '2011-10-28',
  2, '2012-06-24', '2013-04-13'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```

Join `ddi6` and `ddi3` tables; label cases and controls
```{sql, eval = FALSE}
create table ddi7 as 
select t1.person_id, t1.first_date_simvastatin_exposure, t1.last_date_simvastatin_exposure, t2.case_end_date
from (select * from ddi6) t1
left join ddi3 t2 on t1.person_id = t2.person_id
group by t1.person_id, t1.first_date_simvastatin_exposure, t1.last_date_simvastatin_exposure, t2.case_end_date;

--Label cases and controls
alter table ddi7
add column group_label int;

update ddi7
set group_label = 1
where case_end_date is not null;

update ddi7
set group_label = 0
where case_end_date is null;
```
  
Example `ddi7` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~FIRST_DATE_SIMVASTATIN_EXPOSURE, ~LAST_DATE_SIMVASTATIN_EXPOSURE, ~CASE_END_DATE, ~GROUP_LABEL,
  1, '2010-03-21', '2011-10-28', '2011-10-28', 1,
  2, '2012-06-24', '2013-04-13', '<null>', 0
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Calculate 12 months after `first_date_simvastatin_exposure`; calculate 12 months before `case_end_date`
```{sql, eval = FALSE}
create table ddi8 as
select t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure, t1.last_date_simvastatin_exposure, t1.case_end_date, add_months(t1.first_date_simvastatin_exposure, 12) as twelve_months_post_exposure
from (select * from ddi7) t1
group by t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure, t1.last_date_simvastatin_exposure, t1.case_end_date, twelve_months_post_exposure;

alter table ddi8
add column twelve_months_prior_adr date; 

update ddi8 
set twelve_months_prior_adr = add_months(case_end_date, -12)
where case_end_date is not null;
```
  
Example `ddi8` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~FIRST_DATE_SIMVASTATIN_EXPOSURE, ~LAST_DATE_SIMVASTATIN_EXPOSURE, ~CASE_END_DATE, ~TWELVE_MONTHS_POST_EXPOSURE, ~TWELVE_MONTHS_PRIOR_ADR,
  1, 1, '2010-03-21', '2011-10-28', '2011-10-28', '2011-03-21', '2010-10-28', 
  2, 0, '2012-06-24', '2013-04-13', '<null>', '2013-06-24', '<null>'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Define observation period for controls
```{sql, eval = FALSE}
create table ddi9 as 
select t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure as start_date, t1.last_date_simvastatin_exposure, t1.twelve_months_post_exposure
from (select * from ddi8 where group_label = 0) t1
group by t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure, t1.last_date_simvastatin_exposure, t1.twelve_months_post_exposure;

alter table ddi9
add column end_date date; 

update ddi9
set end_date = twelve_months_post_exposure
where last_date_simvastatin_exposure > twelve_months_post_exposure;

update ddi9
set end_date = last_date_simvastatin_exposure
where end_date is null;
```
  
Example `ddi9` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~START_DATE, ~LAST_DATE_SIMVASTATIN_EXPOSURE, ~TWELVE_MONTHS_POST_EXPOSURE, ~END_DATE,
  2, 0, '2012-06-24', '2013-04-13', '2013-06-24', '2013-04-13'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Define observation period for cases
```{sql, eval = FALSE}
create table ddi10 as
select t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure, t1.case_end_date as end_date, t1.twelve_months_prior_adr
from (select * from ddi8 where group_label = 1) t1
group by t1.person_id, t1.group_label, t1.first_date_simvastatin_exposure, end_date, t1.twelve_months_prior_adr;

alter table ddi10 
add column start_date date; 

update ddi10
set start_date = twelve_months_prior_adr 
where first_date_simvastatin_exposure < twelve_months_prior_adr;

update ddi10 
set start_date = first_date_simvastatin_exposure
where start_date is null;
```
  
Example `ddi10` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~FIRST_DATE_SIMVASTATIN_EXPOSURE, ~END_DATE, ~TWELVE_MONTHS_PRIOR_ADR, ~START_DATE,
  1, 1, '2010-03-21', '2011-10-28', '2010-10-28', '2010-10-28' 
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
Merge cases and controls observation period tables
```{sql, eval = FALSE}
create table ddi11 as
(select t1.person_id, t1.group_label, t1.start_date, t1.end_date
from (select * from ddi9) t1
group by t1.person_id, t1.group_label, t1.start_date, t1.end_date)
union 
(select t1.person_id, t1.group_label, t1.start_date, t1.end_date
from (select * from ddi10) t1
group by t1.person_id, t1.group_label, t1.start_date, t1.end_date);
```

Example `ddi11` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~START_DATE, ~END_DATE, 
  1, 1, '2010-10-28', '2011-10-28', 
  2, 0, '2012-06-24', '2013-04-13' 
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```

### Obtain all drugs listed in allergy list for patients in `ddi11` before and on `end_date`
```{sql, eval = FALSE}
create table ddi12 as
select t1.person_id, t1.note_text, min(t1.note_date) as first_adr_date
from (select * from note) t1, (select * from ddi11) t2
where t1.person_id in (t2.person_id)
and lower(note_source_value) like '%allerg%'
group by t1.person_id, t1.note_text; 

create table ddi13 as 
select t1.person_id, t1.note_text, t1.first_adr_date, t2.start_date, t2.end_date
from (select * from ddi12) t1 
left join ddi11 t2 on (t1.person_id = t2.person_id)
group by t1.person_id, t1.note_text, t1.first_adr_date, t2.start_date, t2.end_date;

create table ddi14 as 
select t1.person_id, t1.note_text, t1.first_adr_date, t1.start_date, t1.end_date
from (select * from ddi13) t1
where t1.first_adr_date <= t1.end_date
group by t1.person_id, t1.note_text, t1.first_adr_date, t1.start_date, t1.end_date;
```
  
Example `ddi14` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~NOTE_TEXT, ~FIRST_ADR_DATE, ~START_DATE, ~END_DATE,
  1, 'gemfibrozil', '2011-09-01', '2010-10-28', '2011-10-28',
  2, 'No known allergies', '2012-09-15', '2012-06-24', '2013-04-13'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
### Obtain outpatient drug exposures for patients in `ddi11` during observation period
```{sql, eval = FALSE}
create table ddi15 as 
select t1.person_id, t1.drug_exposure_start_date as date, t1.drug_source_value, t1.drug_concept_id
from (select * from drug_exposure) t1, (select * from ddi11) t2
where t1.person_id in (t2.person_id)
and t1.drug_exposure_start_date >= t2.start_date
and t1.drug_exposure_start_date <= t2.end_date
group by t1.person_id, t1.drug_exposure_start_date, t1.drug_source_value, t1.drug_concept_id;

create table ddi16 as 
select t1.person_id, t2.visit_start_date as outpatient_visit_date, t1.drug_source_value, t1.drug_concept_id
from (select * from ddi15) t1, (select person_id, visit_start_date from visit_occurrence where visit_concept_id = 9202) t2
where t1.person_id = t2.person_id
and t1.date = t2.visit_start_date
group by t1.person_id, t2.visit_start_date, t1.drug_source_value, t1.drug_concept_id;

--collapse `ddi16` to earliest date
create table ddi17 as
select t1.person_id, t1.drug_concept_id
from (select * from ddi16) t1
group by t1.person_id, t1.drug_concept_id;
```

Example `ddi17` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~DRUG_CONCEPT_ID,
  1, 1308877, 
  2, 43014173
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```

### Obtain covariates for patients in `ddi11`
```{sql, eval = FALSE}
create table ddi18 as
select t1.person_id, t1.group_label, t1.start_date, t1.end_date, t2.gender_source_value as gender, date(t2.birth_datetime) as dob, t2.race_source_value as race
from (select * from ddi11) t1, (select * from person) t2
where t1.person_id = t2.person_id
group by t1.person_id, t1.group_label, t1.start_date, t1.end_date, gender, dob, race;
```
  
Example `ddi18` table
```{r echo=FALSE}
tribble(
  ~PERSON_ID, ~GROUP_LABEL, ~START_DATE, ~END_DATE, ~GENDER, ~DOB, ~RACE,
  1, 1, as.Date('2010-10-28'), as.Date('2011-10-28'), 'F', as.Date('1950-05-06'), 'W', 
  2, 0, as.Date('2012-06-24'), as.Date('2013-04-13'), 'M', as.Date('1965-12-12'), 'W'
) %>% 
  DT::datatable(options = list(scrollX = TRUE))
```
  
### Download tables as CSV file for downstream processing and analysis in R
  
Download `ddi14` as CSV file `ddi_adr_data.csv`  
```{sql, eval = FALSE}
select * from ddi14;
```
  
Download `ddi17` as CSV file `ddi_drug_exposures.csv`
```{sql, eval = FALSE}
select * from ddi17;
```
  
Download `ddi18` as CSV file `ddi_covariates.csv`
```{sql, eval = FALSE}
select * from ddi18;
```
